---
title: "Optimizing seqDB function"
author: "Daniel Gil"
date: "September 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###**Summary**
Multiple options were used to optimize *seqDB* function. In the end, the final function is almost (9x) times faster than the original (see last plot \@ref(fig:final)).

###**Description process**
According to Wickham in his book "Advanced R", optimizing code to make it faster is an iterative procedure: <br>
1. Find the biggest bottleneck (the slowest part of your code).<br>
2. Try to eliminate it (you may not succeed but that’s ok).<br>
3. Repeat until your code is “fast enough.”<br>
This report describes the process conducted to optimize the *seqDB* function. <br>
<br>

### Considerations
- To be able to create multiple versions of a function without losing the location of each file, I re-organized the files that comes from the package. Now each function has its own file.
- I choose the example used in Chapter 3 of the paper "Generating Optimal Designs for Discrete Choice Experiments in R: The idefix Package" to measure the improvements in the optimization process.
- The following packages are used here: <br>
  **profvis**: to find bottlenecks in *seqDB* function<br>
  **microbenchmark**: to compare processing time of each implementation<br>
  **Rcpp**: to implement code in C++.<br>
  **RcppArmadillo**: to use built-in armadillo functions in C++.<br>

## 1. Find the biggest bottleneck <br>
The first step is to load all functions from the package that are used in *SeqDB* function. Because all functions were reorganized, each function need to be loaded:
```{r source}
source("seqDB.R")
source("Derr.R")
source("InfoDes.R")
source("DBerrS.R")
source("DerrS.R")
```

Then all packages are loaded and the example is pasted as in the paper:
```{r packages}
# Load packages
library(microbenchmark)
library(profvis)
library(Rcpp)
library(RcppArmadillo)

# Run the code as in the paper.
set.seed(123)
cs <- idefix::Profiles(lvls = c(4, 3, 2), coding = c("E", "E", "E"))

# Specify prior for each respondent
m <- c(0.5, 0.5, 1, -0.3, -0.7, 0.7)
v <- diag(length(m))
ps <- MASS::mvrnorm(n = 10, mu = m, Sigma = v)

# Generate DB optimal design: 8 choice sets with 2 alternatives each
init.des <- idefix::Modfed(cand.set = cs, n.sets = 8, n.alts = 2,
                           alt.cte = c(0, 0), par.draws = ps)$design
#init.des

#   Simulate choice data for the initial design
#   True individual preference parameter
truePREF <- c(0.8, 1, 1.2, -0.4, -0.8, 1.3)

#   Simulate choices on the logit model
#   In this case, for the first five choice sets the second alternative is 
#   chosen, whereas for the last three the first alternative is chosen.
set.seed(123)
y.sim <- idefix::RespondMNL(par = truePREF, des = init.des, n.alts = 2)
#y.sim 

#   Updating prior distribution
set.seed(123)
draws <- idefix::ImpsampMNL(prior.mean = m, prior.covar = v,
                            des = init.des, n.alts = 2, y = y.sim, m = 6)
#draws

#   Selecting optimal choice
#   minimizing DB-error
dr <- draws$sample
w <- draws$weights
set <- SeqDB(des = init.des, cand.set = cs, n.alts = 2,
             par.draws = dr, prior.covar = v, weights = w)
set

# Profiling seqDB function to find bottlenecks
profvis(SeqDB(des = init.des, cand.set = cs, n.alts = 2,
               par.draws = dr, prior.covar = v, weights = w))
```
Multiple things can be noticed: <br>
- The *profvis* result shows that in the *seqDB* function the bottleneck is located in an apply function. This apply calls the *DBerrS* function.<br>
- In *DBerrS* function, the bottleneck is located in an apply function that calls the *DerrS* function. <br>
- In *DerrS* function, the bottleneck is located in the line that calls the function *InfoDes*.<br>
- In *InfoDes* function, there are multiple lines that takes time to process: a rep(seq()) line. The line where the probability is computed. And the line where the information matrix is computed.<br>
<br>
For these reasons, the biggest bottleneck is in the *InfoDes* function.

## 2. Implementation of *InfoDes* in C++ with Rcpp package.
Before implementing the code in C++, the parameters of the function *seqDB* as well as the variables created before the first apply within the function are created in the global enviroment (like an attach).

```{r param}
# Paramaters.
des <- init.des;des # Optimal design (matrix)
cand.set <- cs;cs # All possible treatments (matrix)
n.alts <- 2 # Number of alternatives for each set (numeric)
par.draws <- dr # Draws from the posterior (matrix)
prior.covar <-  v;prior.covar # Prior covariance matrix (Multinormal) (matrix)
weights <- w;weights # Weights from importance sampling algorithm (numeric)

#----
# Initialize.
n.sets <- nrow(des) / n.alts # Number of sets
cte.des <- NULL # 

# If no weights, equal weights.
if (is.null(weights)) {
  weights <- rep(1, nrow(par.draws)) # Each draw has weight = 1
}

# Detect alternative specific constants
des.f <- as.data.frame(des) # matrix to data.frame to use dplyr functions
alt.cte <- dplyr::select(des.f, dplyr::contains(".cte")) #  select variables 
#that have that string
if (ncol(alt.cte) > 0) {
  cte.des <- alt.cte[1:n.alts, ]  # Why until n.alts?? Ans: It seems that takes the constant defined in the Modfed function. All choice sets have the same constant
}

# Handling par.draws.
if (!(is.matrix(par.draws))) {
  par.draws <- matrix(par.draws, nrow = 1) #  Transform draws that are not matrices to vector. Why? Ans: It seems that is done to give the next error
}

# Error par.draws
if (ncol(des) != ncol(par.draws)) {
  stop("Numbers of parameters in par.draws does not match the number of parameters in the design.")
}

# Error identifying model.
if (n.sets < ncol(par.draws)) { # Dont know why this error yet. Check
  stop("Model is unidentified. Increase the number of choice sets or decrease parameters to estimate.")
}

#----
# Starting and initializing values.
i.cov <- solve(prior.covar) # Inverse of prior covariance matrix
d.start <- apply(par.draws, 1, Derr, des = des,  n.alts = n.alts)

#---- 
# Infodes
group <- rep(seq(1, nrow(des) / n.alts, 1), each = n.alts) #Vector to 

# probability of each alternative in each set
u <- des %*% diag(par.draws[1,])  # 
u <- .rowSums(u, m = nrow(des), n = length(par.draws[1,]))  
p <- exp(u) / rep(rowsum(exp(u), group), each = n.alts)
# information matrix
info.des <- crossprod(des * p, des) - crossprod(rowsum( des * p, group))
```

Then, the implementation is coded and sourced in R to make the comparison. To do this, each line of the *InfoDes* function was coded in C++ and tested. <br>
Having done this, a copy of the *seqDB* function is done, called *seqDB2*, with the following changes: <br>
- Calls the function *Derr2* instead of *Derr*, which calls the function *Infodes_cpp* instead of *InfoDes*. <br>
- Calls the function *DBerrS2* instead of *DBerrS*, which calls the function *DerrS2* which calls *Infodes_cpp* instead of *InfoDes*. <br>
<br>
The comparison of the results and time to execute is presented:


```{Rcpp echo=F, cache=T}
// [[Rcpp::depends(RcppArmadillo)]]

# include <RcppArmadillo.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericMatrix InfoDes_cpp(NumericVector par, NumericMatrix des,
                             double n_alts) {
  int i = 0;
  NumericVector group(des.nrow());
  int ind = 0;
  int cont = 1;
  //group <- rep(seq(1, nrow(des) / n.alts, 1), each = n.alts) #Vector to 
  //indicate the choice set
  // ToDo: try to improve => rep function in cpp: 
  //https://stackoverflow.com/questions/28442582/reproducing-r-rep-with-the-times-argument-in-c-and-rcpp
  for (i = 0; i < des.nrow(); i++){
    if (cont <= n_alts){
      group[i] = ind;
      cont++;
    } else{
      ind++;
      group[i] = ind;
      cont = 2;
    }
  }
  
  // probability
  arma::vec par_arma(par.begin(),des.ncol(),false); //Initializing arma vector 
  arma::mat diagonal = arma::diagmat(par_arma);
  
  // Multiplication
  // Create arma objects;
  arma::mat des_arma(des.begin(), des.nrow(), des.ncol(), false);
  arma::mat u = des_arma * par_arma;
  
  // Exponential
  arma::mat u_exp = exp(u);
  
  // Sum for each choice set
  NumericMatrix uexp = wrap(u_exp);
  NumericMatrix rowsum(des.nrow()/n_alts,1);
  cont = 1;
  int index = 0;
  for( i = 0; i < des.nrow(); i++){
    if( cont <= n_alts){
      rowsum(index,0) += u_exp(i,0);
      cont++;
    }else{
    index++;
    rowsum(index,0) += u_exp(i,0);
    cont = 2;
    }
  }
  
  // Repite each value n_alts times;
  NumericMatrix rowsum_rep(des.nrow(),1);
  cont = 1;
  index = 0 ;
  for(i = 0; i < des.nrow(); i++){
    if (cont <= n_alts){
      rowsum_rep[i] = rowsum[index];
      cont++;
    }else{
      index++;
      rowsum_rep[i] = rowsum[index];
      cont=2;
    }
  }
  
  // Probability
  NumericMatrix p(des.nrow(),1);
  for(i = 0; i < des.nrow(); i++){
    p[i] = uexp[i] / rowsum_rep[i];
  }
  
  // information matrix
  //  Crossprod 1
  // des_p = des*p
  NumericMatrix des_p(des.nrow(),des.ncol());
  for(int j = 0; j < des.ncol(); j++){
    for( i = 0; i < des.nrow(); i++){
      des_p(i,j) = des(i,j) * p[i];
    }
  }
  
  // crossprod(des * p, des)
  arma::mat des_p_arma(des_p.begin(), des_p.nrow(), des_p.ncol(), false);
  arma::mat cross_1 = des_p_arma.t() * des_arma;
  
  //  Crossprod 2
  //rowsum( des * p, group)
  NumericMatrix des_p_rowsum(des_p.nrow()/n_alts,des_p.ncol());
  for(int j=0; j < des_p.ncol(); j++){ 
    cont = 1;
    index = 0;
    for( i = 0; i < des_p.nrow(); i++){
      if( cont <= n_alts){
        des_p_rowsum(index,j) += des_p(i,j);
        cont++;
      }else{
        index++;
        des_p_rowsum(index,j) += des_p(i,j);
        cont = 2;
      }
    }
  }
  
  // crossprod(des * p, des)
  arma::mat des_p_rowsum_arma(des_p_rowsum.begin(), des_p_rowsum.nrow(), des_p_rowsum.ncol(), false);
  arma::mat cross_2 = des_p_rowsum_arma.t() * des_p_rowsum_arma;
  
  // Info.des
  //info.des <- crossprod(des * p, des) - crossprod(rowsum( des * p, group))
  arma::mat info_des = cross_1 - cross_2;
  return(wrap(info_des));
}
```

```{r bench_1_2}
#sourceCpp("InfoDes_cpp.cpp")
source("seqDB2.R")
set <- SeqDB(des = init.des, cand.set = cs, n.alts = 2,
             par.draws = dr, prior.covar = v, weights = w)
set2 <- SeqDB2(des = init.des, cand.set = cs, n.alts = 2,
               par.draws = dr, prior.covar = v, weights = w)
set;set2

a <- microbenchmark(seqDB = SeqDB(des = init.des, cand.set = cs, n.alts = 2,
                     par.draws = dr, prior.covar = v, weights = w),
               seqDB2 = SeqDB2(des = init.des, cand.set = cs, n.alts = 2,
                      par.draws = dr, prior.covar = v, weights = w));a
autoplot.microbenchmark(a)
```
The implementation in C++ is almost **six times (6x) faster** than the original function. <br>

## 3. Find the next biggest bottleneck <br>
As it was mentioned before, optimization is an iterative process. So the next step is to profile this new function to find new bottlenecks.<br>
To do so, all elements are removed from the workspace and the example showed above is run again. Then the profiling is analized:

```{r provis_seqDB2}
 profvis(SeqDB2(des = init.des, cand.set = cs, n.alts = 2,
             par.draws = dr, prior.covar = v, weights = w))
```

These results are similar as in the first function:
- The *profvis* result shows that in the *seqDB2* function the bottleneck is located in an apply function. This apply calls the *DBerrS* function.<br>
- In *DBerrS2* function, the bottleneck is located in an apply function that calls the *DerrS2* function. <br>
- In *DerrS2* function, there are two lines, besides *InfoDes*, that takes time to process: the calculation of the determinant of the information matrix, and the append of two matrices.<br>
<br>
For these reasons, the next bottleneck is in the *DerrS* function.

## 4. Implementation of *InfoDes* in C++ with Rcpp package.
As above, before implementing the code in C++, the parameters of the function *seqDB* as well as the variables created before the first apply within the function are created in the global enviroment (like an attach). This time there other variables that are needed.
```{r echo=FALSE}
# Paramaters.
des <- init.des;des # Optimal design (matrix)
cand.set <- cs;cs # All possible treatments (matrix)
n.alts <- 2 # Number of alternatives for each set (numeric)
par.draws <- dr # Draws from the posterior (matrix)
prior.covar <-  v;prior.covar # Prior covariance matrix (Multinormal) (matrix)
weights <- w;weights # Weights from importance sampling algorithm (numeric)
reduce <- TRUE

n.sets <- nrow(des) / n.alts # Number of sets
cte.des <- NULL # 
# Detect alternative specific constants
des.f <- as.data.frame(des)
alt.cte <- dplyr::select(des.f, dplyr::contains(".cte"))
if (ncol(alt.cte) > 0) {
  cte.des <- alt.cte[1:n.alts, ]
}
```

```{r}
# Starting and initializing values.
i.cov <- solve(prior.covar) # Inverse of prior covariance matrix
d.start <- apply(par.draws, 1, Derr2, des = des,  n.alts = n.alts) #  Calculate the D-error for each alternative
db.start <- mean(d.start, na.rm = TRUE)  # Calculates the mean D-error
full.comb <- gtools::combinations(n = nrow(cand.set), r = n.alts, repeats.allowed = !reduce) # Calculates all possible combinations without repetition
n.par <- ncol(par.draws) # Number of parameters

# For each potential set, select best. 
# db.errors <- apply(full.comb, 1, DBerrS2, cand.set, par.draws, des, n.alts, cte.des, i.cov, n.par, weights)

#   DBerrS2 Function
set <- as.matrix(cand.set[as.numeric(full.comb[1,]), ]) # matrix with only the alternatives chosen from the matrix of all combinations (full.comb).
# Add alternative specific constants if necessary
if (!is.null(cte.des)) {
  set <- as.matrix(cbind(cte.des, set))
}
# For each draw calculate D-error.
# d.errors <- apply(par.draws, 1, DerrS2, set, des, n.alts, i.cov, n.par)

#   DerrS2 Function
des.f <- rbind(des, set)  # Append of optimal design with new alternatives
info.d <- InfoDes(par = par.draws[1,], des = des.f, n.alts = n.alts)  # Information matrix for each row
d.error <- det(info.d + i.cov)^(-1 / n.par) # Calculate sequential d-error
```
Then, the implementation is coded and sourced in R to make the comparison. To do this, each line of the *DerrS* function was coded in C++ and tested. <br>
Having done this, a copy of the *seqDB* function is done, called *seqDB3*, with the following changes: <br>
- Calls the function *Derr2* instead of *Derr*, which calls the function *Infodes_cpp* instead of *InfoDes*. <br>
- Calls the function *DBerrS3* instead of *DBerrS*, which calls the function *DerrS_cpp* instead of *DerrS*. <br>
<br>
The comparison of the results and time to execute is presented:

```{Rcpp echo=F, cache=T}
// [[Rcpp::depends(RcppArmadillo)]]

# include <RcppArmadillo.h>

using namespace Rcpp;

// [[Rcpp::export]]
double DerrS_cpp(NumericVector par, NumericMatrix set, NumericMatrix des,
                        double n_alts, NumericMatrix inv_cov, double n_par) {
  //https://github.com/petewerner/misc/wiki/RcppArmadillo-cheatsheet#create
  // des.f <- rbind(des, set)  # Append of optimal design with new alternatives
  arma::mat des_arma(des.begin(), des.nrow(), des.ncol(), false);
  arma::mat set_arma(set.begin(), set.nrow(), set.ncol(), false);
  arma::mat des_f = join_cols(des_arma, set_arma);
  
 // Another way to do the rbind: (It's slower)
 /*NumericMatrix out = no_init_matrix(des.nrow()+set.nrow(), des.ncol());
 for (int j = 0; j < des.nrow()+set.nrow(); j++) {
   if (j < des.nrow()) {
     out(j, _) = des(j, _);
   } else {
     out(j, _) = set(j - des.nrow(), _);
   }
 }*/
 //Rcout << "arma:" << des_f << std::endl;
 //Rcout << "std:" << out << std::endl;
 
  // Call function InfoDes_cpp
  //info.d <- InfoDes(par = par.draws, des = des.f, n.alts = n.alts) 
  //NumericMatrix info_d(des.ncol(),des.ncol());
  Function InfoDes_cpp( "InfoDes_cpp" ) ; 
  NumericMatrix info_d = InfoDes_cpp(par,wrap(des_f),n_alts);
  
  // Calculate determinant
  //d.error <- det(info.d + i.cov)^(-1 / n.par)
  arma::mat info_d_arma(info_d.begin(), info_d.nrow(), info_d.ncol(), false);
  arma::mat inv_cov_arma(inv_cov.begin(), inv_cov.nrow(), inv_cov.ncol(), false);
  arma::mat sum_1 = info_d_arma + inv_cov_arma;
  double det = arma::det(sum_1);
  double d = pow(det,(-1 / n_par));
  //Rcout << "det= "<<det<<std::endl;
  //Rcout << "d= "<<d<<std::endl;
  //double det = pow(det(sum_1),(-1 / n_par);
  return(d);
  //return(wrap(sum_1));
  //return(info_d);
}

 
 
 // [[Rcpp::export]]
 double det_cpp(NumericMatrix set) {
   arma::mat set_arma(set.begin(), set.nrow(), set.ncol(), false);
   return(arma::det(set_arma));
 }
```

```{r bench_1_3}
#sourceCpp("InfoDes_cpp.cpp")
sourceCpp("DerrS_cpp.cpp")
source("seqDB2.R")
source("seqDB3.R")

set <- SeqDB(des = init.des, cand.set = cs, n.alts = 2,
             par.draws = dr, prior.covar = v, weights = w)
set2 <- SeqDB2(des = init.des, cand.set = cs, n.alts = 2,
               par.draws = dr, prior.covar = v, weights = w)
set3 <- SeqDB3(des = init.des, cand.set = cs, n.alts = 2,
               par.draws = dr, prior.covar = v, weights = w)
set;set2;set3

a <- microbenchmark(seqDB = SeqDB(des = init.des, cand.set = cs, n.alts = 2,
                    par.draws = dr, prior.covar = v, weights = w),
              seqDB2 = SeqDB2(des = init.des, cand.set = cs, n.alts = 2,
                     par.draws = dr, prior.covar = v, weights = w),
              seqDB3 = SeqDB3(des = init.des, cand.set = cs, n.alts = 2,
                    par.draws = dr, prior.covar = v, weights = w));a
autoplot.microbenchmark(a)
```

As a result, the implementation of *DerrS* does not improve the processing time with respect to *seqDB2*, which only has *InfoDes* in C++.<br> 
The reason for this finding is that the base function *rbind* is faster that the implementation done in C++.<br>
However the process time of computing the determinant of the information is faster in C++:

```{Rcpp echo=F, cache=T}
// [[Rcpp::depends(RcppArmadillo)]]

# include <RcppArmadillo.h>

using namespace Rcpp;

 // [[Rcpp::export]]
 double det_cpp(NumericMatrix set) {
   arma::mat set_arma(set.begin(), set.nrow(), set.ncol(), false);
   return(arma::det(set_arma));
 }

```

```{r}
# Calculate determinant
#sourceCpp("DerrS_cpp.cpp")
det(info.d + i.cov)^(-1 / n.par) # Calculate sequential d-error
# The determinant in cpp is faster
det_cpp(info.d + i.cov)^(-1 / n.par) # Calculate sequential d-error
a = microbenchmark(R = det(info.d + i.cov)^(-1 / n.par),
               cpp = det_cpp(info.d + i.cov)^(-1 / n.par));a
autoplot.microbenchmark(a)
```

The source code of *rbind* is also done in C++, however it is much longer and complicated because it works for different kinds of R-objects. For this reason, the *rbind* function is kept in the *DerrS* function, and the determinant is computed using the C++ function. <br>

These changes are saved in a new function called *seqDB4* and compared with the previous functions:
```{r bench_1_4}
#sourceCpp("InfoDes_cpp.cpp")
#sourceCpp("DerrS_cpp.cpp")
source("seqDB2.R")
source("seqDB3.R")
source("seqDB4.R")

set <- SeqDB(des = init.des, cand.set = cs, n.alts = 2,
             par.draws = dr, prior.covar = v, weights = w)
set2 <- SeqDB2(des = init.des, cand.set = cs, n.alts = 2,
               par.draws = dr, prior.covar = v, weights = w)
set3 <- SeqDB3(des = init.des, cand.set = cs, n.alts = 2,
               par.draws = dr, prior.covar = v, weights = w)
set4 <- SeqDB4(des = init.des, cand.set = cs, n.alts = 2,
               par.draws = dr, prior.covar = v, weights = w)
set;set2;set3;set4

a <- microbenchmark(seqDB = SeqDB(des = init.des, cand.set = cs, n.alts = 2,
                    par.draws = dr, prior.covar = v, weights = w),
              seqDB2 = SeqDB2(des = init.des, cand.set = cs, n.alts = 2,
                     par.draws = dr, prior.covar = v, weights = w),
              seqDB3 = SeqDB3(des = init.des, cand.set = cs, n.alts = 2,
                    par.draws = dr, prior.covar = v, weights = w),
              seqDB4 = SeqDB4(des = init.des, cand.set = cs, n.alts = 2,
                    par.draws = dr, prior.covar = v, weights = w));a
autoplot.microbenchmark(a)
```
It can be seen, that this new function is again faster than the previous ones.<br>

<br>
Now, given that a function to compute the determinant is created in C++, the function *Derr* is also modified by calling *det_cpp* function instead of *base::det*. 
So a new function with all these changes is created, called *seqDB5* and compared:
```{r bench_1_5}
#sourceCpp("InfoDes_cpp.cpp")
#sourceCpp("DerrS_cpp.cpp")
source("seqDB2.R")
source("seqDB3.R")
source("seqDB4.R")
source("seqDB5.R")

set <- SeqDB(des = init.des, cand.set = cs, n.alts = 2,
             par.draws = dr, prior.covar = v, weights = w)
set2 <- SeqDB2(des = init.des, cand.set = cs, n.alts = 2,
               par.draws = dr, prior.covar = v, weights = w)
set3 <- SeqDB3(des = init.des, cand.set = cs, n.alts = 2,
               par.draws = dr, prior.covar = v, weights = w)
set4 <- SeqDB4(des = init.des, cand.set = cs, n.alts = 2,
               par.draws = dr, prior.covar = v, weights = w)
set5 <- SeqDB5(des = init.des, cand.set = cs, n.alts = 2,
               par.draws = dr, prior.covar = v, weights = w)
set;set2;set3;set4;set5

a = microbenchmark(seqDB1 = SeqDB(des = init.des, cand.set = cs, n.alts = 2,
                       par.draws = dr, prior.covar = v, weights = w),
               seqDB2 = SeqDB2(des = init.des, cand.set = cs, n.alts = 2,
                          par.draws = dr, prior.covar = v, weights = w),
               seqDB3 = SeqDB3(des = init.des, cand.set = cs, n.alts = 2,
                      par.draws = dr, prior.covar = v, weights = w),
               seqDB4 = SeqDB4(des = init.des, cand.set = cs, n.alts = 2,
                      par.draws = dr, prior.covar = v, weights = w),
               seqDB5 = SeqDB5(des = init.des, cand.set = cs, n.alts = 2,
                                par.draws = dr, prior.covar = v, weights = w));a
autoplot.microbenchmark(a)
```
This results suggest that the last function is the faster one.

###**Conclusion**
After having implemented some functions in C++, the process time of original function is optimized in several orders of magnitude (almost 9x times faster).
The modifications are as follows:<br>
- Infodes is totally rewritten in C++ <br>
- A new function to compute the determinant in C++ is implemented <br>

In this sense, *seqDB* function now should call *InfoDes_cpp* and *det_cpp*.
```{r final_plot, fig.cap="final"}
autoplot.microbenchmark(a)
```
